Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-02-10 pkg/disttree/inst/draft/draft.Rnw
finished revising intro (for now)
various clean-ups (removing commented LaTeX and R code)
graphics and label enhancements



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2005 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9888e50ba8f7c219354e60299b6488da4bd3de7c

## define distribution list:
# dist_list_normal
{

  dist_list_normal <- list()

  ddist <-  function(y, eta, log = TRUE, weights = NULL, sum = FALSE) {

    val <- -1/2 * (log(2*pi) + 2*eta[2] + exp(log((y-eta[1])^2) - 2*eta[2]))
    if(!log) val <- exp(val)

    # par <- c(eta[1], exp(eta[2]))
    # val <- dnorm(y, mean = par[1], sd = par[2], log = log)

    if(sum) {
      if(is.null(weights)) weights <- rep.int(1, length(y))
      val <- sum(weights * val, na.rm = TRUE)
    }
    return(val)
  }

  sdist <- function(y, eta, weights = NULL, sum = FALSE) {

    score <- cbind(exp(-2*eta[2]) * (y-eta[1]), 
                   -1 + exp(-2*eta[2] + log((y-eta[1])^2)))

    # par <- c(eta[1], exp(eta[2])) 
    # score <- cbind(1/par[2]^2 * (y-par[1]), 
    #                (-1/par[2] + ((y - par[1])^2)/(par[2]^3)) * exp(eta[2]))

    score <- as.matrix(score)
    colnames(score) <- etanames
    if(sum) {
      if(is.null(weights)) weights <- rep.int(1, length(y))
      # if score == Inf replace score with 1.7e308 because Inf*0 would lead to NaN -> gradient is NaN
      score[score==Inf] = 1.7e308
      score <- colSums(weights * score, na.rm = TRUE)
    }
    return(score)
  }

  hdist <- function(y, eta, weights = NULL) {    
    ny <- length(y)
    if(is.null(weights)) weights <- rep.int(1, ny)

    d2ld.etamu2 <- sum(weights * rep.int(-exp(-2*eta[2]), ny))
    d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-eta[1]) * exp(-2*eta[2]), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
    d2ld.etasigma2 <- sum(weights * (-2)*exp(log((y-eta[1])^2) - 2*eta[2]), na.rm = TRUE)

    # par <- c(eta[1], exp(eta[2]))                           
    # d2ld.etamu2 <- sum(weights * rep.int(-1/par[2]^2, ny))
    # d2ld.etamu.d.etasigma <- sum(weights * (-2)*(y-par[1])/par[2]^2), na.rm = TRUE)          # should be 0 for exact parameters (here: observed hess)
    # d2ld.etasigma2 <- sum(weights * (-2)*(y-par[1])^2/par[2]^2, na.rm = TRUE)

    hess <- matrix(c(d2ld.etamu2, d2ld.etamu.d.etasigma, d2ld.etamu.d.etasigma, d2ld.etasigma2), nrow = 2)
    colnames(hess) <- rownames(hess) <-  etanames

  ## additional functions pdist, qdist, rdist
  pdist <- pnorm
  qdist <- qnorm
  rdist <- rnorm

  startfun <- function(y, weights = NULL){
    if(is.null(weights)) {
      mu <- mean(y)
      sigma <- sqrt(1/length(y) * sum((y - mu)^2))
    } else {
      mu <- weighted.mean(y, weights)
      sigma <- sqrt(1/sum(weights) * sum(weights * (y - mu)^2))
    }
    starteta <- c(mu, log(sigma))
    names(starteta) <- etanames
    return(starteta)
  }

  mle <- TRUE

  dist_list_normal <- list(family.name = "Normal Distribution",
                           ddist = ddist, 
                           sdist = sdist, 
                           hdist = hdist,
                           pdist = pdist,
                           qdist = qdist,
                           rdist = rdist,
                           link = link, 
                           linkfun = linkfun, 
                           linkinv = linkinv, 
                           linkinvdr = linkinvdr,
                           startfun = startfun,
                           mle = mle,
                           gamlssobj = FALSE,
                           censored = FALSE
  )
}

@@

\Address{
  Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"at Innsbruck \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/}, \url{http://retostauffer.org/}, \url{https://eeecon.uibk.ac.at/~zeileis/}\\

  Torsten Hothorn\\
  Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
  Universit\"at Z\"urich \\
  Hirschengraben 84\\
  CH-8001 Z\"urich, Switzerland \\
  E-mail: \email{Torsten.Hothorn@R-project.org}\\
  URL: \url{http://user.math.uzh.ch/hothorn/}\\

@@

\begin{figure}[t!]
\begin{tikzpicture}
\node (a) at (0,0){
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GLM, echo=FALSE, results=hide>>=
nobs <- 200
## GLM
set.seed(7)
x <- c(1:nobs)/nobs
ytrue <- 1+x*1.5
y <- ytrue + rnorm(nobs,0,0.3)
@

<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
\end{center}
\endminipage};
\node (b) at (5,0) 
{
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAM, echo=FALSE, results=hide>>=
## GAM
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3 
y <- ytrue + rnorm(nobs,0,0.3)
@

<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
\end{center}
\endminipage};
\node (c) at (10,0) 
{
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAMLSS, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(2*x)^2)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@

<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
\end{center}
\endminipage};
\node (d) at (0,-2) 
{
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
LM, GLM  
\end{center}
\endminipage};
\node (e) at (5,-2) 
{
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
GAM 
\end{center}
\endminipage};
\node at (10,-2)
{
\minipage{0.28\textwidth}
\begin{center}
\vspace{0.0cm}
GAMLSS 
\end{center}
\endminipage};
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
\end{tikzpicture}
\caption{\label{Devel_parmod}Parametric modeling developments. (Generalized) linear models (left), generalized additive models (middle), generalized additive models for location, scale, and shape (right).}
\end{figure}

@@

\begin{figure}[t!]
\begin{tikzpicture}
\node (a) at (0,0){
\minipage{0.29\textwidth}
\begin{center}
<<motivation_regtree, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- yforest <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
y <- ytrue + rnorm(nobs,0,0.3)
for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
@

<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = ytree, col = pal["forest"], lwd=7)
@
\end{center}
\endminipage};
\node (b) at (5,0) 
{
\minipage{0.29\textwidth}
\begin{center}
<<motivation_randforest, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
@

<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
@
\end{center}
\endminipage};
\node (c) at (10,0) 
{
\minipage{0.29\textwidth}
\begin{center}
<<motivation_distforest, echo=FALSE, results=hide>>=
## distforest
set.seed(4723)
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
xt <- 2*(x-0.5)
var <- exp(-(2*xt)^4)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
x <- c(1:nobs)/nobs
@

<<plot_motivation_distforest, fig=TRUE, echo=FALSE>>=
par(mar = c(2, 0, 2, 0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
points(x, y, col = "slategray", pch = 19)
box(lwd=8)
@
\end{center}
\endminipage};
\node (d) at (0,-2) 
{
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
Regression Tree 
\end{center}
\endminipage};
\node (e) at (5,-2) 
{
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
Random Forest 
\end{center}
\endminipage};
\node at (10,-2)
{
\minipage{0.28\textwidth}
\begin{center}
\vspace{0.0cm}
Distributional Forest 
\end{center}
\endminipage};
\node (g) at (0,-5.4) 
{
\minipage{0.29\textwidth}
\vspace{0.2cm}
%\hspace{0.0cm}
\begin{center}
<<motivation_disttree, echo=FALSE, results=hide>>=
## disttree
set.seed(74)
x <- c(1:nobs)/nobs
var <- ytree/3
y <- ytrue + rnorm(nobs,0,var)
@

<<plot_motivation_disttree, fig=TRUE, echo=FALSE>>=
par(mar = c(2, 0, 2, 0))
plot(x, y, xaxt="n", yaxt="n", ann=FALSE, type = "n")
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
lines(x = x, y = ytree, col = pal["forest"], lwd=7, main = "")
polygon(c(x, rev(x)), c(ytree + var, rev(ytree - var)),
  col = pallight["forest"], border = "transparent")
points(x, y, col = "slategray", pch = 19)
box(lwd=8)
@
\vspace{-0.2cm}%\\
Distributional Tree
\vspace{0.2cm}
\end{center}
\endminipage};
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](7.1,0)--(7.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](0,-2.4)--(0,-3.2);
\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](2.2,-5.4)--(7.9,-2.5);
\end{tikzpicture}
\caption{\label{Devel_treeforest}Tree and forest developments. Regression tree (top left), distributional tree (bottom), random forest (top middle), distributional forest (top right).}
\end{figure}

In summary, the models discussed so far provide a broad and powerful toolset for
parametric distributional fits depending on a specified set of additive linear or
smooth nonlinear terms. A rather different approach to capturing the dependence on
covariates are tree-based models.
\begin{itemize}
  \item \emph{Regression trees} (\citealt{breiman1984classification}) recursively
    split the data into more homogeneous subgroups and can thus capture abrupt shifts
    (Figure~\ref{Devel_treeforest}, top left) and approximate nonlinear functions.
    Furthermore, trees automatically carry out a forward selection of covariates and
    their interactions. 
  \item \emph{Random forests} (\citealt{breiman2001random}) average the predictions of
    an ensemble of trees fitted to resampled versions of the learning data. This
    stabilizes the recursive partitions from individual trees and hence better
    approximates smooth functions (Figure~\ref{Devel_treeforest}, top middle)
\end{itemize}
While classical regression trees and random forests only model the mean of the response
we propose to follow the ideas from (semi-)parametric regression modeling -- as outlined
in Figure~\ref{Devel_parmod} -- and combine tree-based methods with parametric distributional models,
yielding two novel techniques:
\begin{itemize}
  \item \emph{Distributional trees} split the data into more homogenous groups with
    respect to a parametric distribution, thus capturing changes in any distribution
    paramer like location, scale, or shape (Figure~\ref{Devel_treeforest}, bottom left)
  \item \emph{Distributional forests} utilize an ensemble of distributional trees
    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{Devel_treeforest},
    top right).
\end{itemize}

In many modeling situations, especially when the types of effects are not known in advance,
it can be challenging to specify a well fitting model. The novel methodology of distributional 
trees and forests provides a framework that covers a wide range of effects including linear and nonlinear
additive and non-additive effects as well as interactions of covariates. Moreover, these models are
particularly advantageous in case of many covariates with unknown effects or
interactions as the variable selection is done automatically within the model
fitting process. Therefore, no additional effort or knowledge for the
specification of the model is required in advance.
For these reasons, distributional trees and forests provide
valuable features which can contribute to improving models for probabilistic
forecasting. For example in weather forecasting in mountainous regions various influencing factors 
can change rapidly and their effects are  difficult to be represented by a smooth parametric model. 
Additionally, large spacing of the grid applied in numerical weather prediction models over 
complex terrain can lead to abrupt changes within just one cell of the grid.

In the application discussed in this paper precipitation forecasts are made for a mountainous region
in the Alps covering mainly Tyrol in Austria and adjacent areas. Predictions are based on data from 
85 observation stations and 28 years (1985--2012). The total amount of precipitation within 24 hours
is modeled with the assumption of the underlying distribution being a censored normal distribution 
with censoring point 0. As covariates a large number of meteorological quantities are provided by 
ensembles of NWPs. 
Next to distributional forests other methods commonly used in this research area are applied and their 
results are compared. In a first setting only one observation station is taken into consideration.
The models are trained on data from 24 years and probabilistic predictions of the consecutive 4 years 
are validated. A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}. 
For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated 
by a distributional forest are plotted together with the corresponding observations. This illustrates 
that appropriate probabilistic forecasts are provided by this model as they are close to the observed 
values. Beside the expected values the predictions made by distributional models offer a wide range 
of information as they specify the whole distribution and therefore enable further methods of inference 
and analysis. In section \ref{ProbabilisticForecasting} this application will be further discussed and it 
will be shown that the predictions provided by the distributional forest are reasonably well fitting, not 
only for the mean parameter of the distribution but also for its second parameter, the scale parameter, 
and therefore the whole distribution.

%In the field of regression analysis many statistical models have already been developed to model the relationship between a response variable and one or more explanatory variables.\\
%Among the first were linear models (LM) which predict the (conditional) expected value of the response variable as a linear combination of the explanatory variables. This idea was extended to generalized linear models (GLM) which broadened the range of possible distributions of the response variable to the exponential family. (In R these models can be built using the functions \code{lm()} and \code{glm()} provided by the package \pkg{stats}).\\
%Then, in order to also allow for non-linear effects, generalized additive models (GAM) were introduced where smooth functions of the explanatory variables are summed up instead of the linear combination. (In the fitting function \code{gam} types of smoothing terms can be selected.)\\
%However, with these methods only the mean of the distribution of the response variable can be modeled. Generalized additive models for location, scale,and shape
%(GAMLSS) (\citealt{stasinopoulos2005}) made it possible to model each of the parameters of a distribution separately by its own GAM. In that way a whole distribution can be specified. The R-packages \pkg{gamlss} and \pkg{gamlss.dist} offer the corresponding software with a wide variety of distributions.\\
%This development in the field of parametric modeling is illustrated in Figure~\ref{Devel_parmod}.\\

%All these parametric models work very well as long as only additive effects appear in the considered data. A different type of model that can capture non-additive effects as well are regression trees. Here the basic idea is to split the data set into more homogeneous subgroups based on information provided by the explanatory variables and to fit a model/response to each terminal node. One problem that might appear when applying tree algorithms is that they impose steps and abrupt changes on the model, even in situations where effects are rather smooth, as it can be seen in the top left plot of Figure~\ref{Devel_treeforest}. One way how to tackle this problem is to not just consider one tree but an ensemble of trees. This is the idea of random forests. A set of trees is built, each on a slightly different data set which can either be a bootstrap or a subsample of the original data set. Then the results of all trees are combined to get a final model which unlike a tree model can also deal with smooth effects in a reasonable way.\\ 
%In distributional trees and forests we now want to combine all these ideas from different fields of statistical modeling in order to benefit from all their advantages.\\
%\\
%In the past the combination of parametric data models and algorithmic tree models has gained more and more interest since it allows for a more flexible and clear structure. \\
%Algorithms that only fit constants to each final node and therefore only model the mean of the response variable (such as CART, Breiman 1984) might lead to very large trees which complicates analyzing and interpreting. Algorithms such as GUIDE (Loh 2002), CRUISE (Kim and Loh 2001) and LOTUS (Chan and Loh 2004) were among the first to provide trees with parametric models in each terminal node. The MOB algorithm (Zeileis, Hothorn and Hornik 2008), an algorithm for model-based recursive partitioning, extended this idea by fitting parametric models to each node and using the gained information of the fitted model to create further splits in inner nodes. Using this algorithm as a framework to build trees LMs and GLMs can be fitted in the nodes of the trees. However, up to now trees which only model the mean of the response variable in its nodes are still more common and widely used.\\
%Random forests (\citealt{breiman2001random}) applying the CART algorithm to build the trees and cforest applying the ctree algorithm (\citealt{hothorn2015partykit}, \citealt{hothorn2006unbiased}) have used this idea of combining an ensemble of trees to get smoother effects of the covariates after combining or averaging over the parameters of the trees.\\
%\\
%But as mentioned before, until now the above listed development in the field of probabilistic modeling has not yet been fully integrated in the idea of algorithmic tree and forest models. Probabilistic models are rarely used in combination with tree algorithms and more flexible models such as GAMLSSs have not yet been applied as models in the nodes of a tree.\\
%\\
%Therefore, our main goal is to embed the progress made in the field of probabilistic modeling in the idea of regression trees and forests. In particular, more flexible models such as GAMLSSs should be fitted in the nodes of a tree in order to capture location, scale,and shape as well as censoring, tail behavior etc. while non-additive effects or interactions of the explanatory variables can be detected automatically by the splitting algorithm used to build the tree. In that way, no variable selection or predefining of variable interactions is necessary and a whole distribution is specified in each node which leads to a wide range of inference methods that can be applied.\\
%These (future) steps are illustrated in Figure~\ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\

%After explaining the novel methodology in Section~\ref{Methodology} of this paper the focus will then be put on an application in the field of probabilistic forecasting in Section~\ref{ProbabilisticForecasting}. The new method is applied on precipitation data in a mountainous region with a very large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area. In a first setting only one observation station is taken into consideration. The models are trained on data from 24 years and probabilistic predictions of the consecutive 4 years are validated. A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}. For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated by a distributional forest are plotted together with the corresponding observations. This illustrates that well fitting predicted expected values 
%%% well fitting predicted expected values -> appropriate probabilistic forecasts? </TH>
%are provided by this model as they are close to the observed values. Beside the expected values the predictions provided by distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis. In section \ref{ProbabilisticForecasting} this example will be further discussed and it will be shown that the predictions made by the distributional forest are reasonably well fitting, not only for the mean parameter of the distribution but also for its second parameter, the scale parameter, and therefore the whole distribution.

@@

%The data set consists of daily 24h precipitation sums for the month of July from 28 years, starting in 1985 until 2012, both years included. The observations were measured at 95 (?) stations all over Tyrol (and regions close to the frontier). The considered data is provided by .....
%Next to the variable of interest which is the measured total precipitation within 24 hours of each day (measured from 6 UTC to 6 UTC of the following day), numerical forecasts (from ... ) are used as covariates. These forecasts are ensemble predictions consisting of ... individual forecasts based on varying/perturbed initial conditions.\\

@@

%\newpage
\section*{Appendix}

>>>>

@@

\Address{
  Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
  Universit\"at Innsbruck \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at},\\
  \email{Achim.Zeileis@R-project.org} \\
  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
  \url{http://retostauffer.org/}, \url{https://eeecon.uibk.ac.at/~zeileis/}\\

  Torsten Hothorn\\
  Universit\"at Z\"urich \\
  Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
  Hirschengraben 84\\
  CH-8001 Z\"urich, Switzerland \\
  E-mail: \email{Torsten.Hothorn@R-project.org}\\
  URL: \url{http://user.math.uzh.ch/hothorn/}\\

@@

\begin{figure}[t!]
\begin{tikzpicture}
\node (a) at (0,0){
\minipage{0.29\textwidth}
\centering
<<motivation_GLM, echo=FALSE, results=hide>>=
nobs <- 200
## GLM
set.seed(7)
x <- c(1:nobs)/nobs
ytrue <- 1+x*1.5
y <- ytrue + rnorm(nobs,0,0.3)
@

<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
\endminipage};
\node (b) at (5,0) 
{
\minipage{0.29\textwidth}
\centering
<<motivation_GAM, echo=FALSE, results=hide>>=
## GAM
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3 
y <- ytrue + rnorm(nobs,0,0.3)
@

<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
\endminipage};
\node (c) at (10,0) 
{
\minipage{0.29\textwidth}
\centering
<<motivation_GAMLSS, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(2*x)^2)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@

<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
\endminipage};
\node (d) at (0,-2) 
{
\minipage{0.29\textwidth}
\centering
\vspace{0.0cm}
LM, GLM  
\endminipage};
\node (e) at (5,-2) 
{
\minipage{0.29\textwidth}
\centering
\vspace{0.0cm}
GAM 
\endminipage};
\node at (10,-2)
{
\minipage{0.28\textwidth}
\centering
\vspace{0.0cm}
GAMLSS 
\endminipage};
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
\end{tikzpicture}
\caption{\label{Devel_parmod}Parametric modeling developments. (Generalized) linear models (left), generalized additive models (middle), generalized additive models for location, scale, and shape (right).}
\end{figure}

@@

\begin{figure}[t!]
\begin{tikzpicture}
\node (a) at (0,0){
\minipage{0.29\textwidth}
\centering
<<motivation_regtree, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- yforest <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
y <- ytrue + rnorm(nobs,0,0.3)
for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
@

<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = ytree, col = pal["forest"], lwd=7)
@
\endminipage};
\node (b) at (5,0) 
{
\minipage{0.29\textwidth}
\centering
<<motivation_randforest, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
@

<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
@
\endminipage};
\node (c) at (10,0) 
{
\minipage{0.29\textwidth}
\centering
<<motivation_distforest, echo=FALSE, results=hide>>=
## distforest
set.seed(4723)
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
xt <- 2*(x-0.5)
var <- exp(-(2*xt)^4)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
x <- c(1:nobs)/nobs
@

<<plot_motivation_distforest, fig=TRUE, echo=FALSE>>=
par(mar = c(2, 0, 2, 0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
points(x, y, col = "slategray", pch = 19)
box(lwd=8)
@
\endminipage};
\node (d) at (0,-2) 
{
\minipage{0.29\textwidth}
\centering
\vspace{0.0cm}
Regression Tree 
\endminipage};
\node (e) at (5,-2) 
{
\minipage{0.29\textwidth}
\centering
\vspace{0.0cm}
Random Forest 
\endminipage};
\node at (10,-2)
{
\minipage{0.28\textwidth}
\centering
\vspace{0.0cm}
Distributional Forest 
\endminipage};
\node (g) at (0,-5.4) 
{
\minipage{0.29\textwidth}
\vspace{0.2cm}
%\hspace{0.0cm}
\centering
<<motivation_disttree, echo=FALSE, results=hide>>=
## disttree
set.seed(74)
x <- c(1:nobs)/nobs
var <- ytree/3
y <- ytrue + rnorm(nobs,0,var)
@

<<plot_motivation_disttree, fig=TRUE, echo=FALSE>>=
par(mar = c(2, 0, 2, 0))
plot(x, y, xaxt="n", yaxt="n", ann=FALSE, type = "n")
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
lines(x = x, y = ytree, col = pal["forest"], lwd=7, main = "")
polygon(c(x, rev(x)), c(ytree + var, rev(ytree - var)),
  col = pallight["forest"], border = "transparent")
points(x, y, col = "slategray", pch = 19)
box(lwd=8)
@
\vspace{-0.2cm}%\\
Distributional Tree
\vspace{0.2cm}
\endminipage};
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](7.1,0)--(7.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](0,-2.4)--(0,-3.2);
\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](2.2,-5.4)--(7.9,-2.5);
\end{tikzpicture}
\caption{\label{Devel_treeforest}Tree and forest developments.
Regression tree (top left), distributional tree (bottom left), random forest
(top middle), and distributional forest (top right).}
\end{figure}

In summary, the models discussed so far provide a broad and powerful toolset for
parametric distributional fits depending on a specified set of additive linear or
smooth nonlinear terms. A rather different approach to capturing the dependence on
covariates are tree-based models.
\begin{itemize}
  \item \emph{Regression trees} (\citealt{breiman1984classification}) recursively
    split the data into more homogeneous subgroups and can thus capture abrupt shifts
    (Figure~\ref{Devel_treeforest}, top left) and approximate nonlinear functions.
    Furthermore, trees automatically carry out a forward selection of covariates and
    their interactions. 
  \item \emph{Random forests} (\citealt{breiman2001random}) average the predictions of
    an ensemble of trees fitted to resampled versions of the learning data. This
    stabilizes the recursive partitions from individual trees and hence better
    approximates smooth functions (Figure~\ref{Devel_treeforest}, top middle)
\end{itemize}
While classical regression trees and random forests only model the mean of the response
we propose to follow the ideas from GAMLSS modeling -- as outlined
in Figure~\ref{Devel_parmod} -- and combine tree-based methods with parametric distributional models,
yielding two novel techniques:
\begin{itemize}
  \item \emph{Distributional trees} split the data into more homogenous groups with
    respect to a parametric distribution, thus capturing changes in any distribution
    paramer like location, scale, or shape (Figure~\ref{Devel_treeforest}, bottom left)
  \item \emph{Distributional forests} utilize an ensemble of distributional trees
    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{Devel_treeforest},
    top right).
\end{itemize}
In the following, particular focus is given to distributional forests as a method for obtaining
probabilistic forecasts by leveraging the strengths of random forests: the ability to
capture both smooth and abruptly changing functions along with simultaneous selection of
variables and possibly complex interactions. Thus, these properties make the method particularly
appealing in case of many covariates with unknown effects where it would be challenging
to specify a GAMLSS.

In weather forecasting, these properties are appealing in mountainous regions and complex terrain
where small-scale effects are not well-resolved in the NWP outputs. As these are computed at
a coarser scale, there may be abrupt local effects in the meteorological observations within a single
NWP grid cell. To illustrate this in practice, precipitation forecasts are obtained with distributional
forests at 85 meteorological stations in a mountainous region in the Alps, covering mainly Tyrol,
Austria, and adjacent areas. More specifically, a zero-censored Gaussian distribution is
employed to model 24-hour total precipitation so that the zero-censored point mass describes the
probability of no precipitation on a given day (see Figure~\ref{axams_onestation}). Forecasts
for July are established based on data from the same month in the years 1985--2012 along with
a large number of NWP ensemble outputs as covariates. As Figure~\ref{axams_onestation} shows,
the station-wise models yield a full distributional forecast for each day -- here for the same
day (July~24) at one station (Axams) over four years (2009--2012) -- based on the previous 24~years as training data.
The corresponding observations conform reasonably well with the predictions.

@@

@@

\newpage
\begin{appendix}
